---
title: "Module 4 - Starter Code"
author: "David Buckeridge"
date: "January 19, 2021"
output: html_document
---

```{r setup, include=F, message=F}
options(gsubfn.engine = "R")
library("knitr")
library("sqldf")

# Set your working directory as needed.
# opts_knit$set(root.dir = "", fig.align = "center")
```
```{r data, echo=F}
hospital_discharges = read.csv('data/hospital_discharges.csv')
physician_services = read.csv('data/physician_services.csv')
sampled_patients = read.csv('data/sampled_patients.csv')

hospital_discharges$admit = as.Date(hospital_discharges$admit)
hospital_discharges$discharge = as.Date(hospital_discharges$discharge)
physician_services$date = as.Date(physician_services$date)
sampled_patients$dob = as.Date(paste(as.character(sampled_patients$dob), "-01", sep = ""), format = "%Y-%m-%d")
```

## Diabetes Mellitus Case Definition

Popular algorithm (Hux et al, 2002):

* Two physician diabetes diagnosis codes separated by 730 days or less OR 
* One hospitalization diabetes diagnosis code.

| Measure      | Performance |
|:-------------|:------------|
| Sensitivity: | 86%         | 
| Specificity: | 97%         |

Codes:

* ICD-9:250.*
* ICD-10: E10\*-E14\*

## 1. Identify people with at least one hospital admission for diabetes.

Steps

* We take the date of first admission only (i.e., "min(admit)"), so there will be only one row for each person
* When an aggregtion operator is used in the SELECT clause, there must be a GROUP BY 
* Ranges of ICD9 and ICD10 codes are given to account for the change in coding

```{r q1_1}
hospital_diag = sqldf(
  "SELECT anon_id, min(admit) AS diab_date 
   FROM hospital_discharges 
   WHERE (icd_type='ICD-9' AND icd LIKE '250%')
      OR (icd_type='ICD-10' AND (icd LIKE 'E10%' OR icd LIKE 'E11%' OR
                                 icd LIKE 'E12%' OR icd LIKE 'E13%' OR
                                 icd LIKE 'E14%'))
    GROUP BY anon_id")
```

* Identify all hospitalization events for diabetes
* We take all admissions, so there will be multiple rows for a person if they had multiple admissions

```{r q1_2}
hospital_diag_events = sqldf(
  "SELECT anon_id, admit AS diab_date 
   FROM hospital_discharges 
   WHERE (icd_type='ICD-9'  AND icd LIKE '250%')
      OR (icd_type='ICD-10' AND (icd LIKE 'E10%' OR icd LIKE 'E11%' OR
                                 icd LIKE 'E12%' OR icd LIKE 'E13%' OR
                                 icd LIKE 'E14%'))")
```


**Question A -  What proportion of all subjects had at least one admission for diabetes (R)?**

```{r q1a}
# Place any additional code here
```

**Question B - Plot the frequency distribution of the number of hospitalizations per person for diabetes (SQL + R).**
__Hint: This is easy if you perform another SQL query very similar to the one for hospital_diag...__
```{r q1b}
# Place any additional code here
```

**Question C - What proportion of all hospital admissions were for diabetes (R)?**
```{r q1c}
# Place any additional code here
```

**Question D - When was ICD-10 first used for coding hospital discharges (SQL)?**
```{r q1d}
# Place any additional code here
```

**Question E - Compare the proportion of hospital admissions that were for diabetes before and after the switch to ICD-10 (SQL + R).**

__Rates Before...__
```{r q1e1}
hospital_diag_preICD10 = sqldf(
  "SELECT anon_id, discharge
   FROM hospital_discharges
   WHERE icd_type='ICD-9' 
     AND icd LIKE '250%'")

hospital_discharge_count_preICD10 = sqldf(
  "SELECT COUNT(*)
   FROM hospital_discharges
   WHERE icd_type='ICD-9'")[1,1]  

nrow(hospital_diag_preICD10) / hospital_discharge_count_preICD10
```

__Rates after...__

```{r q1e2}
# Place any additional code here
```

## 2 - Identify physician billing events for diabetes

```{r q2}
phys_diab = sqldf(
  "SELECT anon_id, date 
   FROM physician_services 
   WHERE icd LIKE '250%'")

# Physician consulation rate
phys_diab_unique = sqldf(
  "SELECT DISTINCT anon_id
   FROM physician_services
   WHERE icd LIKE '250%'")
```

**Question A - What proportion of all subjects had at least one physician visit for diabetes (R)?**
```{r q2e1}
# Place any additional code here
```

**Question B - Plot the frequency distribution of physician visits per person for diabetes (SQL + R).**
```{r q2e2}
# Place any additional code here
```

**Question C - What proportion of all physician visits were for diabetes (R)?**
```{r q2e3}
# Place any additional code here
```

## 3 - Identify people with two or more billings for diabetes within 730 days

```{r q3}
phys_diag = sqldf(
  "SELECT x.anon_id, (x.date - y.date) as interval, min(x.date) as diab_date
   FROM phys_diab x 
   JOIN phys_diab y ON x.anon_id=y.anon_id 
                   AND x.date > y.date 
                   AND (x.date - y.date <=730) 
   GROUP BY x.anon_id")
```

**Question A - What proportion of subjects had two visits for diabetes within 730 days?**
```{r q3e1}
# Place any additional code here
```

**Question B - Plot the proportion of subjects that would be identified in 3a as a function of the cutoff ranging from 0 to 730 days. Do you think 730 days is a reasonable cut-off? Explain.**
```{r q3e2}
# Place any additional code here
```

## 4 - Join cases detected through physician billing with those detected from hospital discharges.

```{r q4}
both_diag = sqldf(
  "SELECT anon_id, diab_date FROM phys_diag 
   UNION
   SELECT anon_id, diab_date FROM hospital_diag")
                   
diab_dates = sqldf(
  "SELECT anon_id, min(diab_date) as diab_date 
   FROM both_diag 
   GROUP BY anon_id")

```

**Question A - What proportion of subjects met the case definition for diabetes of either one hospital admission or two physician visits within 730 days (R)?**
```{r q4e1}
# Place any additional code here
```

**Question B - Create a Venn diagram (the areas does not need to perfectly reflect the proportions) to illustrate the following: total number of patients, number that are cases due to physician billing, hospital discharges, or both (R).**
```{r q4e2}
# library(VennDiagram)
# Place any additional code here
```

**Question C - Calculate the proportion of subjects that meet the cased definition stratified by age (44 and under, 45 to 64, 65 and over) and sex (male, female) stratified rates. (Hint - you will need to join to the patients table) (SQL + R).**
```{r q4e3}
# Place any additional code here
```

